---
title: "Health Data Visualization Practice"
output: 
  html_document: 
    fig_height: 8
    fig_width: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
setwd("~/Desktop/MIT/dataviz/") # change to your own directory
```

```{r}
library(RPostgreSQL)
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(ggcorrplot)
library(purrr)

# load the PostgreSQL driver
drv <- dbDriver("PostgreSQL")

# create a connection to the postgres database
# set the search path to the mimiciii schema
con <- dbConnect(drv, dbname = "mimic",
                 host = "localhost", port = 5432,
                 user = "DavidSasson")
dbSendQuery(con, 'set search_path to mimiciii')

# test connection
# dbExistsTable(con, "patients")

# show a list of tables
dbListTables(con)
```

This snippet makes several assumptions with respect to the way the database is set up.

- `dbname`: It assumes the name of your database is the same as your local username (i.e., the result of `whoami`). If this is not the case, you should pass a different value indicating the name of your database (e.g., `dbname='mimic'`).
- `host`: It assumes the database is hosted locally on your machine (i.e., on `localhost`).
- `port`: It assumes Postgres is listening on the default port, `5432`.
- `user`: It assumes the current user has access to the database.
- `password`: It assumes no password is required for this user, usually because Postgres is using `peer` authentication.

********


# Data Exploration with ggplot2

```{r}
admins = dbGetQuery(con, "select * from admissions")

str(admins)
```

```{r}
x <- ggplot(admins)

x + geom_bar(aes(religion)) + coord_flip() + theme_minimal()
```


```{r}
x + geom_bar(aes(insurance)) + facet_wrap(~ admission_type) + theme_fivethirtyeight()
```

```{r}
x + geom_bar(aes(ethnicity)) + coord_flip() + theme_tufte()
```

```{r}
x + geom_count(aes(x = insurance, y = ethnicity), alpha=0.3, col="red") + 
  scale_size_area() +
  theme_classic() 
```

```{r}
demo.los <- as.numeric(admins$dischtime - admins$admittime)

ggplot() + 
  geom_histogram(aes(demo.los), alpha=0.6, fill=4, bins = 30) + 
  ggtitle("Length of Stay in ICU") 
```

```{r}
x + geom_bar(aes(discharge_location)) + coord_flip() + ggtitle("Discharge Location")
```



```{r}
# keep only numeric columns 
nums <- admins %>% keep(is.numeric)

ggcorrplot(cor(nums))

pairs(nums, pch =19, lower.panel=panel.smooth, cex = 0.1, upper.panel = NULL)

```



```{r}
dbListTables(con)
```



```{r}
stays = dbGetQuery(con, "select * from icustays")

str(stays)
```


```{r}
t <- ggplot(stays)

# keep only numeric columns 
nums <- stays %>% keep(is.numeric)

ggcorrplot(cor(nums))

# plot histogram of length of stay in the ICU
t + geom_histogram(aes(los), binwidth = 1, fill=I("#9ebcda"), col=I("#FFFFFF")) + 
  xlim(c(0,20)) + 
  ggtitle("Length of stay in the ICU") + 
  xlab("Length of stay, days")
```





##### Optional Break

Anything you guys are intersted in exploring in MIMIC? If so, let's do it together! I promise you'll see me make mistakes every step of the way, but I think we have enough brainpower in this room to find something cool. 

```{r}
dbListTables(con)
```


