---
title: "Health Data Visualization Practice"
output: 
  html_document: 
    fig_height: 8
    fig_width: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
setwd("~/Desktop/MIT/dataviz/") # change to your own directory
```

```{r}
library(RPostgreSQL)
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(ggcorrplot)
library(purrr)

# load the PostgreSQL driver
drv <- dbDriver("PostgreSQL")

# create a connection to the postgres database
# set the search path to the mimiciii schema
con <- dbConnect(drv, dbname = "mimic",
                 host = "localhost", port = 5432,
                 user = "DavidSasson")
dbSendQuery(con, 'set search_path to mimiciii')

# test connection
# dbExistsTable(con, "patients")

# show a list of tables
dbListTables(con)
```

This snippet makes several assumptions with respect to the way the database is set up.

- `dbname`: It assumes the name of your database is the same as your local username (i.e., the result of `whoami`). If this is not the case, you should pass a different value indicating the name of your database (e.g., `dbname='mimic'`).
- `host`: It assumes the database is hosted locally on your machine (i.e., on `localhost`).
- `port`: It assumes Postgres is listening on the default port, `5432`.
- `user`: It assumes the current user has access to the database.
- `password`: It assumes no password is required for this user, usually because Postgres is using `peer` authentication.

********


# Data Exploration with ggplot2

```{r}
admins = dbGetQuery(con, "select * from admissions")

str(admins)
```

```{r}
x <- ggplot(admins)

x + geom_bar(aes(religion)) + coord_flip() + theme_minimal()

x + geom_bar(aes(insurance)) + facet_wrap(~ admission_type) + theme_fivethirtyeight()

x + geom_bar(aes(ethnicity)) + coord_flip() + theme_tufte()

x + geom_count(aes(x = insurance, y = ethnicity), alpha=0.3, col="red") + 
  scale_size_area() +
  theme_classic() 

demo.los <- as.numeric(admins$dischtime - admins$admittime)

ggplot() + 
  geom_histogram(aes(demo.los), alpha=0.6, fill=4, bins = 30) + 
  ggtitle("Length of Stay in ICU") 

# keep only numeric columns 
nums <- admins %>% keep(is.numeric)

ggcorrplot(cor(nums))

pairs(nums, pch =19, lower.panel=panel.smooth, cex = 0.1, upper.panel = NULL)

```

Full disclosure--I put together this presentaion on my flight over here yesterday :) below is some scratch that didn't make it into the pretty graphs above. I'm including it to show you what my thought process is and what real data exploration looks like. It sure as hell ain't coherent, but you'll always end up with an interesting answer if you keep digging through the data. _Bonus points_ to whoever makes cool vizzes with these *Bonus points mean nothing, but I will really like you as a person* 

```{r}
# look at time difference. why is this straight line? Maybe look at boxplot of diffs for each subject
# x + geom_point(aes(admittime, deathtime), alpha=0.3)

# This takes too long but should be investigated. ~ 15,000 diagnosis codes?
# need to find which 15 are most common and shop
# x + geom_bar(aes(diagnosis)) + coord_flip()
# 
# unique(admins$diagnosis)
# [1000] "CORONARY ARTERY DISEASE\\CORONARY ARTERY BYPASS GRAFT; ? OFF PUMP/SDA"                              
# [ reached getOption("max.print") -- omitted 14692 entries ]
```


```{r}
dbListTables(con)
```



```{r}
stays = dbGetQuery(con, "select * from icustays")

str(stays)
```


```{r}
t <- ggplot(stays)

# keep only numeric columns 
nums <- stays %>% keep(is.numeric)

ggcorrplot(cor(nums))

# plot histogram of length of stay in the ICU
t + geom_histogram(aes(los), binwidth = 1, fill=I("#9ebcda"), col=I("#FFFFFF")) + 
  xlim(c(0,20)) + 
  ggtitle("Length of stay in the ICU") + 
  xlab("Length of stay, days")
```





##### Optional Break

Anything you guys are intersted in exploring in MIMIC? If so, let's do it together! I promise you'll see me make mistakes every step of the way, but I think we have enough brainpower in this room to find something cool. 

```{r}
dbListTables(con)
```


